Liveness-Based Garbage Collection for Lazy Languages

### 2015-05-18 paper/paper.tex
Intro

Functional  programming  languages  depend  on  garbage  collector  to
efficiently  reclaim  unused  memory   allocated  by  programs  either
explicitly (for example using constructors) or implicitly (for example
for      closures).       However,      empirical      studies      on
Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
Java~\cite{shaham02estimating}  programs   have  shown   that  garbage
collectors leave uncollected a lot of memory objects that are not live
(here {\em live}ness  means the object can potentially be  used by the
program at  a later  stage).  Due  to lack of  a suitable  analysis to
distinguish  reachable  objects  from live  objects.  current  garbage
collectors conservatively approximate the  liveness of heap objects by
their reachability from a set of memory locations called the {\em root
  set\/} resulting in unnecessary memory usage.

Languages having lazy  semantics for evaluation put  further demand on
memory as they  require heap allocated closure to be  carried from the
point of occurrence  of an expression to the point  of (first) real use
of  that expression.  Strictness analysis~\cite{some-refs}  statically
determines  the  patterns  of  future usage  of  an  expression,  thus
allowing some  of the  closures to  be evaluated at  the point  of the
occurrence of the corresponding expression.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% MOTIVATING EXAMPLE
\newcommand{\nilfigure}
{\scalebox{0.75}{
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,0)(1,1)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
\putnode{start}{origin}{0}{0}{}
\putnode{stop}{origin}{10}{10}{}
\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
\end{pspicture}
}}

@@

\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
%  \end{center}
%%%%}\color{Myblue} end
\end{figure*}
We propose the  use of liveness analysis of heap  cells and strictness
analysis  of expressions  to  reduce the  memory  requirement of  lazy
functional  programming languages.   We modify  garbage collectors  to
preserve  only {\em  live} data---a  subset of  {\em reachable  data}.
This results in an increase  in the garbage reclaimed and consequently
in fewer  garbage collections.  Further, we eagerly evaluate closures
that are guaranteed to be evaluated in the future (strict closures).

The main contribution  of this paper is an uniform  framework for {\em
  liveness} and {\em strictness}  analysis to enable the optimizations
mentioned  above. Using  a central  notion of  demand, we  formulate a
context-sensitive liveness  analysis for structured data  and prove it
correct.  We then  present strictness as a dual of  liveness. We use a
0-CFA-like conservative approximation to  annotate each allocation and
function-call program point with  a finite-state automaton---which the
garbage-collector inspects to curtail reachability during marking. The
result of  the strictness analysis,  being necessarily finite,  can be
directly used to insert code to evaluate closures.

We previously proposed an interprocedural liveness analysis to improve
garbage collection first order eager subset of Scheme~\cite{cc-paper}.
The current paper extends the work to handle lazy evaluation semantics
for the same language.

Experiments  confirm  the  expected  performance benefits.  The  early
evaluation   of   closures  decreases   the   memory   demands  of   a
program. Further,  liveness-based collection results in  a increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections, a decrease  in the memory size required  to run programs,
and  reduced  overall  garbage  collection  time  for  a  majority  of
programs.

@@

\subsubsection{Organization of the paper}
Section~\ref{sec:defs} gives
the syntax and semantics of
the language used to illustrate our analysis along with basic concepts
and    notations.
% Section~\ref{sec:operational}   describes    the
% operational semantics  of the  language.
Liveness analysis is described in
Section~\ref{sec:liveness} followed by a sketch of a correctness
proof relative to a non-standard semantics.
Section~\ref{sec:computing} shows how to encode liveness as
finite-state automata.
Section~\ref{sec:experiments}   reports   experimental   results and
Section~\ref{sec:lgc-always-better} proves that a liveness based
collector can never do more garbage collections than a reachability
based collector. 
% and Section~\ref{sec:conclusion} concludes the paper.

@@

The body of the program is the expression denoted by $e_\mainpgm$; for
analysis purposes it is convenient to regard $e_\mainpgm$ as part of a
function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
We write  $\pi\!:\!e$ to  associate the label  $\pi$ (not part  of the
language syntax) with the program point just before expression
$e$.

@@

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the  future use of variables, demands  represent the future
use of the  values of expressions. The demand on  an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\RETURN~x)$.  Assume  that  the  context  of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves). 
\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\RETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\RETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots  \IN
	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\RETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%} 
    \end{boxedminipage}

@@

    \begin{boxedminipage}{.46\textwidth}

        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[~ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
              $~~~~~~~~~~~~~~\;\;
                \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
                \rbrace~\rbrack$ \\
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $~[
                  ~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, $\\
$ ~~~\pi_3 \mapsto \lbrace ep_2
                  \rbrace $, $ \pi_4 \mapsto \lbrace ep_2
                  \rbrace$, $\pi_5 \mapsto \lbrace  ep_3 \rbrace$, \\
                $~~~\pi_6 \mapsto \lbrace ep_3 \rbrace,  \pi_7 \mapsto
                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3 \rbrace~\rbrack$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}$, $\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}$, $\ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
                  $\pi_8 \mapsto \lbrace ep_4 \rbrace,~\psi_4 \mapsto \lbrace ep_4 \rbrace~\rbrack$  
                \end{tabular}
        \end{tabular}

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across expressions.  Note that,  unlike eager languages, the syntactic
occurrence  of an expression  in a  program denotes  the point  of 
creation of its closure, not necessarily its evaluation.
%% To   deal  with   this,   we  introduce   an   analysis  which   is
%%flow-insensitive but path-sensitive\comment{Checked the terminology,
%%path-sensitive means certain properties  holding along a path only,
%%http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
To see the  consequences of this, consider an analysis  of the body of
$\length$ with a demand $\sigma$.  Clearly, this is also the demand on
the return value $z$ giving the liveness at the program point $\pi_8$
as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated backwards
as in traditional liveness analysis.

%%%%%%%%%%%%%%%%%%%%%%

% Push this to a later point

%% Unlike eager languages, a variable may be bound to an expression
%%   that is not fully evaluated. i.e. a closure.

%% To  deal with the  first issue,  we apply  the results  of liveness
%% analysis only to fully evaluated data.  If the marking phase during
%% garbage collection leads to a closure, and the closure is not dead,
%% we  mark  the  entire   reachable  memory  under  the  closure  for
%% collection.   A  better scheme  would  be  to  garbage collect  the
%% closure based on  the liveness at the point  of its creation.  This
%% is sound because our  language is statically scoped.  However, this
%% is  work  in progress,  and  the  method  described in  this  paper
%% switches over to reachability while collecting closures.

%% %%%%%%%%%%%%%%%%%%%%%%%%

%% \renewcommand{\arraystretch}{1}{
%% 	  \begin{uprogram}
%% 	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
%% 	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
%% 	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
%%           \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
%%           $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
%% 	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
%%           $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
%% 	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
%% 	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
%%           \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
%% 	\end{uprogram}}
%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \medskip

@@

Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the  liveness
analysis computes two maps:  the function $\mathcal{L}$ computes a map
from a  e-path to a  liveness environment arising  out of all  uses of
variables  along   the  path.  We  call   this  map  $\cal{M}_{\mathit
  f}^{\sigma}$ (or  $\cal{M}_{\mathit f}$ when the  demand $\sigma$ is
clear from  the context).  A  second map $\cal{P}_{\mathit f}$  maps a
program point to  the set of e-paths sharing the  point.  If a program
point happens  to fall on more  than one such path,  then the liveness
environment of  the program  point is the  variable-wise union  of the
liveness  environments  of  the  individual  paths.   In  the  example
program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
ep_3\rbrace$,  the liveness  at $\pi_1$  is  ${\cal M_{\length}}(ep_1)
\cup {\cal  M_{\length}}(ep_2)\cup {\cal M_{\length}}(ep_3)$.   In the
sequel   we   shall  describe   how   to   obtain  $\cal{M}$   through
$\mathcal{L}$.  The  formulation of  $\cal{P}$ is easy  and we  do not
elaborate it any further.

{\color{red}  \hrule \hrule \hrule  \medskip }

@@

evaluation  of  $x$  and  hence  a new  map  ${\cal  M}=[\psi  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
expression $(\SIF~\psi:x~e_1~e_2)$ is  a pointwise-union of the ${\cal
  M}$-maps of $e_1$  and $e_2$. In addition, since  the condition also
triggers an evaluation of $x$, the map $[\psi \mapsto \{x.\epsilon\}]$
is created  and added to the  union.  Also notice  a consequences of
laziness---the  entire  expression  including  the  condition  is  not
evaluated if  the demand  on it is  $\emptyset$.  This results  in the
empty map denoted by $\emptyset_{\cal{M}}$.

@@

\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\

@@

        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\

@@

 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\

@@

        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   

@@

 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}

@@

%======================================================================================
%% \mycomment{In an eager language, \pw\ would not have been live after $\pi_7$, the
%% point of  its last  use. In a  lazy language, however,  the definition
%% $\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
%% $(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
%% takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
%% record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
%% every program  point from $\pi_8$ up  to the beginning  of the program
%% (since  we do  not kill  liveness).  We  thus regard  the body  of the
%% function as  a tree and consider paths  from the root of  this tree to
%% program  points  which  trigger  a evaluation---\SIF\  conditions  and
%% \RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
%% program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
%% $ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
%%  and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.

>>>>

Functional  programming  languages  depend  on  garbage  collector  to
efficiently  reclaim  unused   memory  allocated  by  programs  either
explicitly  (for  example using  constructors)  or implicitly  (during
creation    of    closures).     However,   empirical    studies    on
Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
Java~\cite{shaham02estimating}   programs  have  shown   that  garbage
collectors leave uncollected a lot of memory objects that are not live
(here {\em live}ness  means the object can potentially  be used by the
program at  a later  stage).  Due  to lack of  a suitable  analysis to
distinguish  reachable  objects  from  live objects,  current  garbage
collectors conservatively approximate the  liveness of heap objects by
their reachability from a set of memory locations called the {\em root
  set\/}. This results in unnecessary memory retention.

Programs written in languages having lazy semantics for evaluation put
an further demands on  memory as they require heap allocated
closures to be  carried from the point of  occurrence of an expression
to the point of (first) real use of that expression.
%% This   may   be   mitigated  by   strictness
%% analysis~\cite{some-refs} that  can statically determine  the patterns
%% of future usage  of an expression, thus allowing  some of the closures
%% to be  evaluated at the point  of the occurrence  of the corresponding
%% expression.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% MOTIVATING EXAMPLE
\newcommand{\nilfigure}
{\scalebox{0.75}{
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,0)(1,1)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
\putnode{start}{origin}{0}{0}{}
\putnode{stop}{origin}{10}{10}{}
\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
\end{pspicture}
}}

@@

\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
%  \end{center}
%%%%}\color{Myblue} end
\end{figure*}
We propose the  use of liveness analysis of heap  cells and strictness
analysis  of expressions  to  reduce the  memory  requirement of  lazy
functional  programming languages.   We modify  garbage collectors  to
preserve  only {\em  live} data---a  subset of  {\em reachable  data}.
This results in an increase  in the garbage reclaimed and consequently
in fewer  garbage collections.  
Using  a central notion  of demand,  we formulate  a context-sensitive
liveness  analysis  and  prove   it  correct.   We  use  a  0-CFA-like
conservative   approximation   to   annotate   each   allocation   and
function-call program point  with a finite-state automaton---which the
garbage-collector inspects to curtail reachability during marking.

While this idea has been shown to be effective~\cite{asati14lgc} for a
first-order  eager  language,  a  straightforward  extension  of  this
earlier  technique  is  not   possible  in  lazy  languages,  where  a
heap-allocated   object  may  either   be  a   evaluated  data   or  a
closure. Since liveness is brought about by evaluation of closures and
given the  uncertainty in lazy  languages regarding the  program point
where this  evaluation takes  place, this introduces  complications in
the analysis itself.  Further,  since we want to extend liveness-based
collection to closures, we  track closures  to their
points of creation, from where we obtain their liveness. Since this is
done  at  runtime,  this  requires considerable  modification  to  the
earlier GC scheme.

%% We  previously  proposed an  interprocedural  liveness analysis  to
%% improve   garbage   collection   first   order  eager   subset   of
%% Scheme~\cite{asati14lgc}.   The current paper  extends the  work to
%% handle lazy evaluation semantics for the same language.

Experiments  confirm  the expected  performance  benefits.  The  early
evaluation   of   closures  decreases   the   memory   demands  of   a
program. Further,  liveness-based collection results in  a increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections,  a  (some quantification)  decrease  in  the memory  size
required to run programs,  and reduced overall garbage collection time
for a majority of programs.

@@

\subsubsection{Organization of the paper}
Section~\ref{sec:defs} gives
the syntax and semantics of
the language used to illustrate our analysis along with basic concepts
and    notations.
% Section~\ref{sec:operational} describes the operational semantics of
% the language.
Liveness analysis is  described in Section~\ref{sec:liveness} followed
by  a  sketch  of  a  correctness proof  relative  to  a  non-standard
semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
as   finite-state  automata.    Section~\ref{sec:experiments}  reports
experimental  results  and Section~\ref{sec:lgc-always-better}  proves
that a liveness based collector  can never do more garbage collections
than a reachability based collector.
% and Section~\ref{sec:conclusion} concludes the paper.

@@

The body  of a function ${\mathit  f}$ will be  denoted as $e_{\mathit
  f}$. for  analysis purposes it is convenient  to regard $e_\mainpgm$
as part of a function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$
as in  C\@.  We  write $\pi\!:\!e$ to  associate the label  $\pi$ (not
part  of the  language  syntax)  with the  program  point just  before
expression $e$.

@@

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the  future use of variables, demands  represent the future
use of the  values of expressions. The demand on  an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\RETURN~x)$.  Assume  that  the  context  of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves). 
\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{boxedminipage}{.45\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\RETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\RETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots  \IN
	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\RETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%} 
    \end{boxedminipage}

@@

    \begin{boxedminipage}{.51\textwidth}

        \small
        \begin{tabular}{lc}
        $\mathcal{L}(e_{\length},\sigma,\Lfonly) = $ 
            \begin{tabular}{lc}            
              $[~ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
              $~~~~~~~~~~~~~~\;\;
                \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
                \rbrace~\rbrack$ \\
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $~[
                  ~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, $\\
$ ~~~\pi_3 \mapsto \lbrace ep_2
                  \rbrace $, $ \pi_4 \mapsto \lbrace ep_2
                  \rbrace$, $\pi_5 \mapsto \lbrace  ep_3 \rbrace$, \\
                $~~~\pi_6 \mapsto \lbrace ep_3 \rbrace,  \pi_7 \mapsto
                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3 \rbrace~\rbrack$
              \end{tabular}\\
              $\mathcal{L}(e_{\main},\sigma,\Lfonly) = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}$, $\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}$, $\ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
                  $\pi_8 \mapsto \lbrace ep_4 \rbrace,~\psi_4 \mapsto \lbrace ep_4 \rbrace~\rbrack$  
                \end{tabular}
        \end{tabular}

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across expressions.  Note that,  unlike eager languages, the syntactic
occurrence  of an expression  in a  program denotes  the point  of 
creation of its closure, not necessarily its evaluation.
To see the  consequences of this, consider an analysis  of the body of
$\length$ with a demand $\sigma$.  Clearly, this is also the demand on
the return value $z$ giving the liveness at the program point $\pi_8$
as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated backwards
as in traditional liveness analysis.

@@

Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
analysis  computes  two   maps:  $\mathcal{L}(e_f,  \sigma,  \Lfonly)$
computes a  map that takes a  e-path in ${\mathit f}$  and returns the
liveness environment  arising out of  all uses of variables  along the
e-path.  A  second map $\cal{P}_{\mathit  f}$ maps any program  point in
$\mathit f$ to
the set of  e-paths sharing the point.  If a  program point happens to
fall on  more than one such  e-path, then the  liveness environment of
the  program  point  is   the  variable-wise  union  of  the  liveness
environments of the individual e-paths.  In the example program, since
${\mathcal P}_{\length}(\pi_1) = \lbrace ep_1, ep_2, ep_3\rbrace$, the
liveness at $\pi_1$ is ${\cal  M}(ep_1) \cup {\cal M}(ep_2) \cup {\cal
  M}(ep_3)$, where ${\cal M} = \mathcal{L}(e_f, \sigma, \Lfonly)$.  In
the  sequel we shall  only define  $\mathcal{L}$.  The  formulation of
$\cal{P}$ is easy and we do not elaborate it any further.

@@

evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union  of the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$,  the map  $[\mathit{ep} \mapsto  \{x.\epsilon\}]$ is  created and
added  to the  union.  Also  notice a  consequences  of laziness---the
entire  expression including  the condition  is not  evaluated  if the
demand on it is $\emptyset$.  This results in the empty map, which, by
abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.

@@

\mathcal{L}((\RETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep} \mapsto
x.\sigma\rbrack, \mbox{ where $\mathit{ep}$ is a
  new e-path terminating with $\psi$}\\

@@

        \lbrack\lbrace \mathit{ep} \mapsto  \{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\

@@

&&\mbox{where $\mathit{ep}$ is a new e-path
  terminating with $\psi$}\\

@@

        \lambda\;\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
\mbox{where}~\mathcal{M} =
\mathcal{L}(e,\sigma,\Lfonly),~\Lv \in \mathcal{M}(ep),~\mbox{and}~f~\mbox{is the new function}\\
&& \makebox[0mm]{\hspace*{7cm}   

@@

 \makebox[0mm]{\hspace*{8cm} where
     $x_1\, \ldots\, x_n$ are the variables in $s$}

@@

 The liveness  of {\LET} has  an elegant formulation.   Recollect that
 our language permits lazy {\LET}s whose definitions can be recursive,
 i.e.  in the expression $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in
 $s$.   We  model  the  definition  $x  \leftarrow~s$  as  a  possibly
 recursive function (named $\mathit{f}$ in the rule) which captures an
 arbitrary number of ``unrollings'' of the definition\footnote{The $*$
   in the definition of ${\mathit f}$ represents  non-deterministic choice,
   but can be treated as a constant for purposes of analysis.}.  We pass to the
 demand transformer $\Lfonly_{\mathit f}$  of this function the demand
 arising out  of the  liveness of  $x$ at the  beginning of  $e$. This
 gives the liveness of the variables of $s$.
